\section{Motivation}
\label{sec:motivation}

Historically, many systems were designed for computer systems and architectures where I/O dominates
performance. However, modern systems with multi-core architectures, larger DRAM capacities,
advanced instruction sets, and other hardware accelerants like vector operations (SIMD), which allow us to
perform the same operation on multiple data items simultaneously, have significantly altered this
landscape. Hence, we must reconsider how we implement modern systems.

\subsection{Importance for Database Systems}
In-memory database systems, which store data directly in the main memory rather than on disk,
provide significantly faster data access and processing due to reduced latency.
Therefore, they are no longer I/O bound and, need high intra-operator
parallelism. To fully utilize the multi-core architecture and other hardware features such as cache
locality and SIMD instructions for higher data parallelism should be considered to achieve maximum
performance.

\subsection{Efficient Join Implementation}
In this thesis we want to look at one common database operator: the join \cite{DBLP:conf/cloud/BlanasP13}.
The join operator is a fundamental component of every database system.
In recent years, the difference in performance between the sort-merge and radix-hash join has been
the subject of ongoing debate. Kim et al. \cite{10.14778/1687553.1687564} projected that Sort-Merge Join would outperform hash-based
alternatives with 512-bit SIMD. Albutiu et al. \cite{MPSM} reinforced this claim with recent results reporting that
their implementation of sort-merge join is superior to that of hash joins (without
leveraging SIMD). Balkesen et al. \cite{Balkesen} experimentally show contradicting results by implementing 
optimized versions for sort-merge and radix-hash join, showing that their implementation of
radix-hash join is still superior.

\subsection{Open-Source Implementation}

Despite ongoing research, public implementations of join algorithms optimized for modern hardware
are hard to find. Most existing implementations are proprietary or 
experimental\footnote{Implementation by Balkesen et al.: \url{https://archive-systems.ethz.ch/node/334}}, limiting their
accessibility and usefulness to the research community and database developers. An open-source,
state-of-the-art implementation of a sort-merge join optimized for different architectures would
help address this need. Such an implementation serves as a valuable baseline for researchers
looking to evaluate or improve upon existing methods, and it also contributes to advancing database
system design by providing a solid foundation for future innovation.


% Description of the scientific context (including the classification in the literature, projects, ...) and of the concrete embedding.

