\section{Related Work}
\label{sec:related-work}

% Description of related work in literature or other projects (theses, master projects, etc.).

Many papers describe how SIMD-accelerated sorting can be done efficiently on modern multi-core architectures. 
Chhugani et al. \cite{10.14778/1454159.1454171} describe the concepts 
needed for efficient (SSE) SIMD sorting for both single- and multi-core execution, including sorting
networks, bitonic- and odd-even merge networks, and how to deal with memory bandwidth limitations 
for large problem sizes through multiway merging. There are other ideas like MergePath \cite{MergePath}
for merging only a few very large sublists in parallel and SIMD accelerated.
Kim et al. \cite{10.14778/1687553.1687564} implemented a sort-merge join using SSE intrinsics using these same 
concepts, projecting performance for wider SIMD widths that would outperform hash joins. Albutiu 
et al. \cite{MPSM} present MPSM, a sort-merge join implementation designed for modern multi-core and
multi-socket NUMA processors using their custom sorting routine without SIMD. They did an
experimental evaluation on a 32-core (4 socket) system, concluding that their sort-merge join
implementation is faster than the respective hash join implementation of Blanas et al. \cite{10.1145/1989323.1989328}. 
Recent studies show that parallel radix-hash join has the best overall performance \cite{6544839}.

Therefore,
Balkesen et al. \cite{Balkesen} experimentally studied the performance of in-memory, parallel,
multi-core join, and NUMA-aware algorithms, focusing on sort-merge and radix-hash join.
They claim to provide the fastest in-memory join processing algorithms using sorting and hashing,
and that sort-merge join gets more comparable in performance to radix-hash join with very large 
input sizes. Still, they conclude that the radix-hash join exceeds the sort-merge join for 256-bit
SIMD. None of the papers mentioned above take advantage of 512-bit SIMD. The hash join operator
implemented in Hyrise is based on \cite{6544839} and \cite{Balkesen}, allowing for good comparison
between a sort-merge join operator inside the Hyrise system. Other in-memory databases like DuckDB 
also utilize parallel radix-hash joins.

There is research on SIMD
sorting using AVX-512 \cite{Watkins, 8855628}. Still, to our knowledge, no literature 
exists on implementing sort-merge join with the same optimizations and concepts like
multiway merging for AVX-512. The much more extensive instruction set of AVX-512, with its gather/scatter intrinsics, improved
masking, and new instruction for data manipulation like compress and expand, might allow for
further improvements, besides just offering a wider vector width, by creatively using these new
instructions.
